Future Cosmic Microwave Background delensing with galaxies surveys.

### 2017-05-22 Notes/DES_rec_phi_paper.tex
-mtext

\documentclass[reprint,superscriptaddress,amsmath,amssymb,floatfix,aps,prd]{revtex4-1}
\usepackage{bm}
\usepackage{epsfig}
\usepackage{natbib}
\usepackage{graphicx,epsfig}
\usepackage{hyperref}
\usepackage{ifthen}
\usepackage{xstring}
% \usepackage{graphicx}\usepackage[top=2.5 cm, bottom=2.5 cm, left=1.05in, right=1.05in]{geometry}
\usepackage[applemac]{inputenc}
\usepackage{amsmath,amssymb}
\usepackage{color}
\usepackage{epstopdf}
%\usepackage[caption=false]{subcaption}
%\epstopdfDeclareGraphicsRule{.gif}{png}{.png}{convert gif:#1 png:\OutputFile}
% \AppendGraphicsExtensions{.gif}
% notes to make comment is non-zero if
\newcommand{\al}[1]{\begin{align} #1 \end{align}}

@@

\def\l{\ell}
% For various journals
%\newcommand{\apj}{ApJ}
\newcommand{\physrep}{Physics Reports}
\newcommand{\jcap}{JCAP}
\newcommand{\apjl}{ApJL}
\newcommand{\apjs}{ApJS}
\newcommand{\aap}{A\&A}
\newcommand{\mnras}{MNRAS}
\newcommand{\physrev}{Phys. Rev.}
\newcommand{\physrevlett}{Phys. Rev. Lett.}
\newcommand*\aj{AJ}
%\newcommand*\prd{Phys.~Rev.~D}

@@

\title{Future Cosmic Microwave Background delensing with galaxies surveys.}         % Enter your title between curly braces
\author{A. Manzotti}        % Enter your name between curly braces
\email{manzotti.alessandro@gmail.com}
\affiliation{Kavli Institute for Cosmological Physics, University of Chicago, 5640 South Ellis Avenue, Chicago, IL 60637, USA}
\affiliation{Department of Astronomy and Astrophysics, University of Chicago, 5640 South Ellis Avenue, Chicago, IL 60637, USA}
\date{\today}

        % Enter your date or \today between curly braces
\begin{abstract}
\begin{itemize}
\item Worry : they might ask for internal bias\
item Possible TODO : temperature $N_{eff}$

\end{itemize}

The measurement of the CMB polarization is a promising experimental dataset to test the inflationary paradigm and to probe the physics of the early universe. A particular component, the so-called B-modes is indeed a direct signature of the presence of gravitational waves in the early universe. However improving the level of noise is not enough. This is even truer if the aim is to not only detect the amplitude of gravitational waves but also the shape of their spectrum to test for example inflation consistency relations.
Removing the lensing component from the measurement of CMB B-modes will be important to constrain the amplitude of the primordial gravitational wave contribution to the signal.
Here we discuss the role of current and future large scale structure surveys in improving the reconstruction of the lensing potential that lenses the CMB photons.
We find that..
\end{abstract}

@@

\section{Introduction}
\label{sec:intro}
Inflation Physics with CMB

Current constraints

Now limiting by delensing

Done what are the prospects/

This paper is organized as follow:

\section{Gravitational lensing B-mode}
\label{sec:th}

As for the CMB photons temperature, the intensity map of photons on the sky, also the Q and U mode decomposition of their polarization is remapped by lensing as:
\be
Q(\hat{\mathbf{n}}) = Q_{\mathrm{unlensed}}(\hat{\mathbf{n}}+\nabla\phi);~~
U(\hat{\mathbf{n}}) = U_{\mathrm{unlensed}}(\hat{\mathbf{n}}+\nabla\phi)
\ee
where $\nabla\phi$ is the deflection angle directly related to the lensing potential potential integrated along the line of sight $\phi$.
The CMB polarization is usually decompose into odd-parity Fourier modes E and B. As shown in \cite{} this for symmetry reason the B-modes configuration is mainly sourced by tensor perturbations.
However this promising sign of primordial gravitational waves is partially obscured by a secondary mechanism that can produce B-modes in the late universe. Indeed gravitational interaction with large scale structure generates CMB B-modes by distorting primordial E-modes.

As a first approximation, given a convergence field $\kappa= -\frac{1}{2}\nabla^2\phi $ the B mode resulting from the lensing of primordial E mode is:
\be
B^{\mathrm{lens}}(\bl) =  \int \frac{d^2 \bl'}{(2 \pi)^2} W(\bl,\bl') E(\ublu) \kappa(\bl - \bl')
\label{eqn:blens}
\ee
where different modes contributes with a different weight given by:
\be
W(\bl,\bl') = \frac{2 \bl' \cdot (\bl-\bl')}{|\bl-\bl'|^2} \sin(2\varphi_{\bl,\bl'}),
\ee

As usual we define the power spectrum as:
\be
\langle B^{\mathrm{lens}}(\bl) ~ B^{\mathrm{lens}^*}(\tilde{\bl}) \rangle \equiv (2\pi)^2 \delta^D(\bl - \tilde{\bl}) C^{BB,\mathrm{lens}}_l
\ee

@@

The full B-mode power spectrum measured on the sky is given by a possible primordial component $C_l^{BB,r}$ together with the lensing $C_l^{BB,\mathrm{lens}}$ contribution and instrumental noise $N_l^{BB}$:
\be
C_l^{BB,\mathrm{full}} = C_l^{BB,r} + C_l^{BB,\mathrm{lens}} + N_l^{BB}.
\ee
Given the current constraints, the lensing component is a significant source of B-modes that, for GW searches purposes correspond to a white noise source of roughly $5 \mu K$-arcmin. This means that it is not only bigger than the biggest allowed GW contribution at scales bigger than several degrees but it is comparable with the decreasing level of instrumental noise. 
For this reason, it is very important to characterize it and eventually remove it from the data.
This can be done for example using \refeq{blens} (or its real-space analog) given a measurement of the E-mode field and the lensing potential $\phi$.
While E is measured directly, we ca estimate $\phi$ using "tracers" of the dark matter distribution that creates the potential.

If we have a large scale structure field I($\nver$) that traces the lensing potential responsible for the lensing of the CMB the template of the lensing B mode on the sky is a weighted convolution:
\be
\hat{B}^{\mathrm{lens}}(\bl) = \int \frac{d^2 \bl'}{(2 \pi)^2} W(\bl,\bl') f(\bl,\bl') E^N(\ublu) I(\bl - \bl')
\ee
where $f(\bl,\bl')$ can be determined minimizing the difference with the true $B^{\mathrm{lens}}(\bl)$ defined in \refeq{blens}.

@@

The weights $f(\bl,\bl')$ so that the residual lensing B mode power is minimized are:
\be
f(\bl,\bl') = \left(\frac{C^{EE}_{{l'}}}{C^{EE}_{{l'}}+N^{EE}_{{l'}}}\right)  \frac{C^{\kappa I}_{|\bl-\bl'|}}{C^{II}_{|\bl-\bl'|} }
\ee
Notice that the first term consists in the usual inverse variance filter applied to the measured E-mode and the second minimize the difference between the reconstructed $\phi$ and the CMB lensing potential.

With this choice of $f(\bl,\bl')$ we finally have that the residual power is:
\beqn
C_{{l}}^{BB,\mathrm{res}} &=&   \int \frac{d^2 \bl'}{(2 \pi)^2}  W^2 (\bl,\bl')
 C^{EE}_{l'} C^{\kappa \kappa}_{|\bl-\bl'|}  \\
&\times&  \left[1 - \left(\frac{C^{EE}_{{l'}} }{C^{EE}_{{l'}}+N^{EE}_{{l'}}}\right) \rho^2_{|\bl-\bl'|} \right] \nonumber
\eeqn
with
\be
\rho_l= \frac{\cl^{\kappa I}}{\sqrt{\cl^{\kappa \kappa} \cl^{I I}}}.
\ee

@@

Let's now assume that we have n different tracers of the gravitational potentials $I_{i}$ with $i\in \{1,..,n\}$. It can be shown that the optimal way to combine them to estimate $\phi$ or, in other word, maximizing the correlation factor $\rho$ is:
\beqn
I 		&=&	\sum_{i}c^{i}I^{i} \nonumber \\
c_{i} 	&=& (C^{-1})_{ij}C^{\kappa I^{j}}
\label{eqn:combined}
\eeqn
where C is the covariance matrix of the LSS tracers.
The ``effective'' correlation of these combined tracers with gravitational lensing is:
\be
\rho^{2} = \sum_{i,j}\frac{C^{\kappa i}~(C^{-1})_{ij}~C^{\kappa j}}{C^{\kappa\kappa}}.
\label{eqn:rho-combined}
\ee

Note that gain we have in adding a new tracer is not only proportional to its correlation with the CMB lensing but it also depends on how much it is correlated with the already used set of tracers.
\reffig{kernel} show the different kernels as a function of redshift computed using the models and parameters described in the following sections.
Galaxies clustering surveys can only reconstruct the low-z portion of the lensing kernel as can be seen from the LSST DES and DESI curves.
However given the significant low noise of this measurement they can still help the delensing of the CMB.
Furthermore, the low-redshift tail of the lensing kernel is also the one affected by large scale structures non-linearities. They can indeed be extremely helpful if future CMB surveys will try to remove that contribution instead of attempting to model it.
On the other end, CIB and 21 cm surveys are probing to high redshift structures and independently from the model assumed they show a fairly good overlap with the CMB lensing kernel.

@@

\begin{figure}[htbp]
\begin{center}
\includegraphics[scale=1.]{../images/compare_kernel.pdf}
\caption{Comparison of the different kernels used in this analysis. The bigger the overlap wit the CMB lensing kernel the better the reconstruction of the lensing potential will be. This will lead to a better delensing. }
\label{fig:kernel}

@@

\section{Tracers Model}
\label{sec:data}
In this section we briefly describe how the power spectra of \refsec{th} are computed and what it is assumed for the different tracers considered in this work.

A 2D projected field $\delta^{i}$ that traces dark matter overdensities can be seen as a 3D field projected along the line of site as:

\begin{equation}
\delta^{i}(\nver) = \int_0^{\infty} dz\, W^{i}(z)\delta(\chi(z)\nver,z).
\label{eqn:wkappa}
\end{equation}
where $\delta(\chi(z)\nver,z)$ correspond to the dark matter overdesity field at a comoving distance $\chi(z)$ at redshift $z$ in the angular direction $\nver$.
Using the Limber approximation \cite{limber53} we can compute the power spectra of 2 fields $i,j$ as:
\begin{equation}
C_{\ell}^{ij}= \int_0^{\infty} \frac{dz}{c} \frac{H(z)}{\chi(z)^2}\, W^{i}(z)W^{j}(z)P(k,z).
\label{eqn:wkappa}{}
\end{equation}

In this equation, $H(z)$ is the Hubble factor at redshift $z$, $c$ is the speed of light, $\chi(z)$ the comoving distance $P(k,z)$ is the matter power spectrum evaluated at wavenumber $k = \ell/\chi(z)$ at redshift $z$. Furthermore $W^{i}(z)$ is the kernel function of the field i.
These theoretical quantities have been computed using CAMB and HALOFIT.

\subsection{CMB lensing potential}
\label{sec:kappaCMB}
CMB photons interact gravitationally with low-redshift large scale structure.
The deflection angle is given by $\mathbf{d}(\nver) = \nabla\phi(\nver)$, where $\nabla$ is the the two-dimensional gradient on the sphere. Because the lensing potential is an integrated measure of the projected gravitational potential, taking the two-dimensional Laplacian of the lensing potential we can define the lensing convergence $\kappa(\nver) = -\case{1}{2}\nabla^2\phi(\nver)$, which depends on the projected matter overdensity $\delta$
% \citep{bart01}:
The lensing kernel $W^{\kappa}$ is
%
$\chi(z)$ is the comoving distance to redshift $z$, $\chi_*$ is the comoving distance to the last-scattering surface at $z_*\simeq 1090$
\begin{equation}
W^{\kappa}(z) = \frac{3\Omega_{\rm m}}{2c}\frac{H_0^2}{H(z)}(1+z)\chi(z)\frac{\chi_*-\chi(z)}{\chi_*},
\end{equation}
%
where $\Omega_{\rm m}$ and $H_0$ are the present day values of the Hubble and matter density parameters, respectively.

The CMB lensing potential is \textit{the} field that we need to reconstruct in order to reverse the effect of large scale structure and delens the CMB.
However the lensing potential can also be reconstructed from the CMB itself. In that case it can be see as a noisy tracer of the true field. While the CMB internal reconstruction fiels as the same kernel as the true one $W^{\kappa}(z)$.
Given the instrumental noise level and the beam, we can compute the reconstruction noise $N_l^{\kappa \kappa}$, so that its power spectrum is
\beq
\cl^{\kappa_\mathrm{rec} \kappa_\mathrm{rec}} = \cl^{\kappa \kappa } +N^{\kappa \kappa}_l
\eeq
In this work the level of noise is computed assuming an iterative quadratic estimator is used in the reconstruction.

\begin{figure}[htbp]
\begin{center}
\includegraphics[scale=0.9]{../images/cmb_internal.pdf}
\caption{\textbf{CMB alone:} Correlation factor between CMB internal reconstructed potential and the true lensing potential for different CMB experiments and corresponding noise reconstructions. }
\end{center}
\label{fig:corr-internal}

@@

The galaxy overdensity $g(\nver)$ in a given direction on the sky is also expressed as a line of sight integral of the matter overdensity:
\begin{equation}
g(\nver) = \int_0^{z_*} dz\, W^{g}(z)\delta(\chi(z)\nver,z),
\end{equation}
where the kernel is
\begin{equation}
\label{eqn:wg}
\begin{split}
W^{g}(z) &= \frac{b(z)\frac{dN}{dz}}{\Bigl(\int dz'\,\frac{dN}{dz'}\Bigr)}.
\end{split}
\end{equation}
Here $\frac{dN}{dz}$ is the number of galaxies as a function of redshift observed by the survey while $b(z)$ is the galaxy bias that connect at different redshift the amplitude of galaxies overdensity to the one of the dark matter.
When computing the auto-spectrum of a galaxies density the shot noise term needs to be taken into account. This is done by adding a constant term to the power spectrum equal to the inverse of the number of galaxies per steradians.

Different galaxy surveys in this work are fully characterized by their b(z), $\frac{dN}{dz}$ and the observed galaxies density.
We test the delensing efficiency modeling current survey like WISE or DES as well as future galaxy probes like DESI and LSST together with 21 cm measurement like SKA.

WISE redshift distribution is taken from \cite{yan:2013} (see Fig. 4 therein). 
To compute the noise term we assume that the available scale after masking WISE should be around $f_{sky}=0.44$ with 50 million galaxies \cite{ferraro:2015}.

DES is modeled after the DES Science Verification data public release.

For DESI we used the $\frac{dN}{dz}$ in Tab. 2.3 of the DESI Technical Design Report.

LSST is modeled following \cite{schaan:2016} as $\frac{dN}{dz}\propto z^{\alpha}\exp^{-(z/z_{0})^{\beta}}$ with $\alpha = 1.27$, $\beta = 1.02$, and $z_{0}= 0.5$. Furthermore we assume a density of 26 galaxies per arcmin squared.

We model both the SKA redshift distribution and bias survey following \cite{namikawa:2016a}.

\subsection{Cosmic infrared Background}
\label{sec:cib}
The definition of the kernel of the CIB is more complex and model dependent.
Following \cite{Addison:2011se}, we model the CIB power directly as $C_{\ell}^{{\rm CIB}\mbox{-}{\rm CIB}} = 3500 (l/3000)^{-1.25} {\rm Jy^2 / sr}$.
This model provides an accurate fit for the power of the \textit{Herschel} $500\,\mu$ map used in this work.
For the cross-spectra with the CMB lensing or other galaxies tracers $C_{\ell}^{{\rm CIB}\mbox{-}j}$,we use the single-SED model of~\cite{hall10}.
This places the peak of the CIB emissivity at redshift $z_c = 2$ with a broad redshift kernel of width $\sigma_z = 2$.
This model is rescaled to agree with the results of
~\cite{Holder:2013hqu} and \cite{planck2013XVIII} by choosing the corresponding linear bias parameter.
Other multi-frequency CIB models are available \citep[e.g.,][]{Bethermin:2013nza}; however, given the level of noise, we are relatively insensitive to this choice.
With these assumptions, depending on angular scale, $45-65\%$ of the CIB is correlated with the CMB lensing potential, as shown in \reffig{corrnow}.
% \subsection{Weak Lensing}
\cite{hall:2010,sherwin:2015,ade:2014,hanson:2013,szapudi:2001,0004-637X-567-1-2,planck-collaboration:2014,planck-collaboration:2014a,planck-collaboration:2011,plank-collaboration:2014,boulanger:1996,lewis:2006}

>>>>

\documentclass[reprint,superscriptaddress,amsmath,amssymb,floatfix,aps,prd,nofootinbib]{revtex4-1}
\usepackage{bm}
\usepackage{epsfig}
\usepackage{natbib}
\usepackage{graphicx,epsfig}
\usepackage{hyperref}
\usepackage{ifthen}
\usepackage{xstring}
% \usepackage{graphicx}\usepackage[top=2.5 cm, bottom=2.5 cm, left=1.05in, right=1.05in]{geometry}
\usepackage[applemac]{inputenc}
\usepackage{amsmath,amssymb}
\usepackage{color}
\usepackage{epstopdf}
%\usepackage[caption=false]{subcaption}
%\epstopdfDeclareGraphicsRule{.gif}{png}{.png}{convert gif:#1 png:\OutputFile}
% \AppendGraphicsExtensions{.gif}
% notes to make comment is non-zero if
\newcommand{\al}[1]{\begin{align} #1 \end{align}}

@@

\def\l{\ell}
% For various journals
%\newcommand{\apj}{ApJ}
\newcommand{\physrep}{Physics Reports}
\newcommand{\jcap}{JCAP}
\newcommand{\apjl}{ApJL}
\def\araa{\ref@jnl{ARA\&A}}
\newcommand{\apjs}{ApJS}
\newcommand{\aap}{A\&A}
\newcommand{\mnras}{MNRAS}
\newcommand{\physrev}{Phys. Rev.}
\newcommand{\physrevlett}{Phys. Rev. Lett.}
\newcommand*\aj{AJ}
%\newcommand*\prd{Phys.~Rev.~D}

@@

\title{Future Cosmic Microwave Background delensing with galaxies surveys.}         % Enter your title between curly braces
\author{A. Manzotti}        % Enter your name between curly braces
\email{manzotti.alessandro@gmail.com}
\affiliation{Kavli Institute for Cosmological Physics, University of Chicago}
\affiliation{Department of Astronomy and Astrophysics, University of Chicago, 5640 South Ellis Avenue, Chicago, IL 60637, USA}
\date{\today}

        % Enter your date or \today between curly braces
\begin{abstract}
%\begin{itemize}
%\item Worry : they might ask for internal bias\
%item Possible TODO : temperature $N_{eff}$
%
%\end{itemize}
%

The measurement of the cosmic microwave background (CMB) polarization is a promising experimental dataset to test the inflationary paradigm and to probe the physics of the early universe. 
A particular component, the so-called B-modes is indeed a direct signature of the presence of gravitational waves in the early universe. However improving the level of noise is not enough. 
This is even truer if the aim is to not only detect the amplitude of gravitational waves but also the shape of their spectrum to test for example inflation consistency relations.
Removing the lensing component from the measurement of CMB B-modes will be important to constrain the amplitude of the primordial gravitational wave contribution to the signal.
Here we discuss the role of current and future large scale structure surveys in improving the reconstruction of the lensing potential that lenses the CMB photons and how this reflects into an improved delensing efficiency.
We quantify this by the improvement due to delensing on the constraints on the inflationary tensor perturbations amplitude and shape ($r$ and $n_{t}$)
We find that..
\end{abstract}

@@

\section{Introduction}
\label{sec:intro}
In the standard cosmological paradigm, the universe underwent a period of near-exponential expansion in its early phase;
this period is called ``cosmic inflation.''

Inflation generically predicts a stochastic background of gravitational waves \cite[see e.g.][for a review]{kamionkowski15}.
These primordial gravitational waves in the early universe imprint a unique signature on the polarized anisotropies of the CMB.
CMB polarization is usually decomposed fields into even-parity (divergence) and odd-parity (curl) components, referred to as ``E'' and ``B'' modes.
In the standard scenario, the PGW background is the only source of B-mode polarization at the epoch of recombination.

The amplitude of this primordial B-mode component is parametrized by the ratio of the amplitudes of the primordial tensor and scalar perturbations, $r$. The shape of this component, i.e. the amplitude as a function of the angular scales is encoded in the tensor tilt $n_{t}$.
Most of the inflationary models directly relate the value of r to the energy scale of inflation.
CMB B modes provide the cleanest known observational window into the primordial gravitational waves background.
Measuring or constraint the value of $r$ and $n_t$ will put tight constraint on the allowed models of inflation.

Indeed improving the CMB B-modes measurement is a major objective of current and future CMB experiments.
Even if inflationary B-modes have not been detected yet, a possibly natural value $r \gtrsim 10^{-3}$, which should be reachable given the near future level of noise expected in CMB experiments \citep{kamionkowski15}.

However reducing the level of noise will not be enough to reach this goal.
Unfortunately the observed B modes are not solely due primordial gravitational waves but also by late time effects. In this work we will focus on gravitational interactions with LSS.
Lensing shears the CMB polarization pattern, producing ``lensing B modes'' from CMB E modes \citep{zaldarriaga98}.
This expected component has now been measured both in cross-correlation with LSS \citep{hanson13,polarbear2014c,vanengelen14b,planck2015XV} and directly from CMB data alone \citep{polarbear2014b,bicep2a,keisler15}.

This component acts as a source of confusion  for searches of the PGW background.
Indeed, current experiment are already limited by the contamination from lensing B modes since the instrument noise is below the lensing B mode $rms$ fluctuations \citep{bk14}.
The optimal way to approach this source of noise is to build the specific realization of lensing B modes on the sky can be and used it to clean the data in a process called ``delensing''.

Delensing has been studied for many years \cite{knox2002, kesden2002, seljak2003, simard:2015,sherwin15,smith:2012}.
Furthermore recently it has been performed on CMB temperature data using the cosmic infrared background in \cite{larsen:2016}, on CMB temperature and polarization data using internal CMB reconstruction \cite{ carron17}. Finally the highest B-mode delensing efficiency has been achieved with SPT and Herschel data in  where $~28\%$ of the lensing power has been removed.
\vspace{0.5cm}

In this paper we

\vspace{0.5cm}

This paper is organized as follow: in \refsec{th} we describe the lensing B-mode components and the residual power after delensing with tracers of the lensing potential. We describe the LSS tracers used in this analysis in \refsec{model}. \refsec{for} is the main results of this work: the improvement to inflationary parameters due to delensing with CMB and LSS. We conclude in \refsec{concl}.

\section{Gravitational lensing B-mode and delensing}
\label{sec:th}

As for the CMB photons temperature, the intensity map of photons on the sky, also the Q and U mode decomposition of their polarization is remapped by lensing as:
\be
Q(\hat{\mathbf{n}}) = Q_{\mathrm{unlensed}}(\hat{\mathbf{n}}+\nabla\phi);~~
U(\hat{\mathbf{n}}) = U_{\mathrm{unlensed}}(\hat{\mathbf{n}}+\nabla\phi)
\ee
where the deflection angle is the gradient of lensing potential potential integrated along the line of sight $\nabla\phi$.
The CMB polarization is usually decomposed into odd-parity Fourier modes E and B. 
As shown in \cite{} this for symmetry reason tensor perturbations are the main source of the B-modes configuration. For this reason, B-modes are a promising signature of early universe tensor perturbations.
However, this promising sign of primordial gravitational waves is partially obscured by a secondary mechanism that can produce B-modes also in the late universe. Indeed the gravitational interaction with large scale structure generates CMB B-modes by distorting primordial E-modes.

As a first approximation, given a convergence field $\kappa= -\frac{1}{2}\nabla^2\phi $ the B mode resulting from the lensing of primordial E mode is:
\be
B^{\mathrm{lens}}(\bl) =  \int \frac{d^2 \bl'}{(2 \pi)^2} W(\bl,\bl') E(\ublu) \kappa(\bl - \bl')
\label{eqn:blens}
\ee
where different modes contributes with a different weight given by:
\be
W(\bl,\bl') = \frac{2 \bl' \cdot (\bl-\bl')}{|\bl-\bl'|^2} \sin(2\varphi_{\bl,\bl'}).
\ee

As usual we define the power spectrum as:
\be
\langle B^{\mathrm{lens}}(\bl) ~ B^{\mathrm{lens}^*}(\tilde{\bl}) \rangle \equiv (2\pi)^2 \delta^D(\bl - \tilde{\bl}) C^{BB,\mathrm{lens}}_l .
\ee

@@

The full B-mode power spectrum measured on the sky is given by a possible primordial component $C_l^{BB,r}$ together with the lensing $C_l^{BB,\mathrm{lens}}$ contribution and the instrumental noise $N_l^{BB}$:
\be
C_l^{BB,\mathrm{full}} = C_l^{BB,r} + C_l^{BB,\mathrm{lens}} + N_l^{BB}.
\ee
Given the current constraints, the lensing component is a significant source of B-modes that, for GW searches purposes correspond to a white noise source of roughly $5 \mu K$-arcmin. This means that it is not only bigger than the biggest allowed GW contribution at scales bigger than several degrees but it is comparable with the decreasing level of instrumental noise. 
For this reason, it is critical to characterize and eventually remove it from the data.
To do so, we need to build a template \refeq{blens} (or its real-space analog) of the actual lensing modes given a measurement of the E-mode field and the lensing potential $\phi$.
While E is measured directly, we ca estimate $\phi$ using "tracers" of the dark matter distribution that creates the potential.

We will now show how much the delensing efficiency is related to the fidelity of the lensing tracers.
If we have a large scale structure field I($\nver$) that traces the lensing potential responsible for the lensing of the CMB we can build a template of the lensing B mode on the sky by a weighted convolution:
\be
\hat{B}^{\mathrm{lens}}(\bl) = \int \frac{d^2 \bl'}{(2 \pi)^2} W(\bl,\bl') f(\bl,\bl') E^N(\ublu) I(\bl - \bl')
\ee
where $f(\bl,\bl')$ can be determined byminimizing the difference with the true $B^{\mathrm{lens}}(\bl)$ defined in \refeq{blens}.

@@

The weights $f(\bl,\bl')$ so that the residual lensing B mode power is minimized are:
\be
f(\bl,\bl') = \left(\frac{C^{EE}_{{l'}}}{C^{EE}_{{l'}}+N^{EE}_{{l'}}}\right)  \frac{C^{\kappa I}_{|\bl-\bl'|}}{C^{II}_{|\bl-\bl'|} }
\ee
Notice that the first term consists in an inverse variance filter applied to the measured E-mode and the second minimize the difference between the reconstructed $\phi$ and the CMB lensing potential.

With this choice of $f(\bl,\bl')$ we finally have that the residual power is:
\beqn
\label{eqn:Bres}
C_{{l}}^{BB,\mathrm{res}} &=&   \int \frac{d^2 \bl'}{(2 \pi)^2}  W^2 (\bl,\bl')
 C^{EE}_{l'} C^{\kappa \kappa}_{|\bl-\bl'|}  \\
&\times&  \left[1 - \left(\frac{C^{EE}_{{l'}} }{C^{EE}_{{l'}}+N^{EE}_{{l'}}}\right) \rho^2_{|\bl-\bl'|} \right] \nonumber
\eeqn
with
\be
\rho_l= \frac{\cl^{\kappa I}}{\sqrt{\cl^{\kappa \kappa} \cl^{I I}}}.
\ee

@@

Let's now assume that we have n different tracers of the gravitational potentials $I_{i}$ with $i\in \{1,..,n\}$. We can optimally combine them to estimate $\phi$ or, in other word, maximizing the correlation factor $\rho$ with:
\beqn
I 		&=&	\sum_{i}c^{i}I^{i} \nonumber \\
c_{i} 	&=& (C^{-1})_{ij}C^{\kappa I^{j}}
\label{eqn:combined}
\eeqn
where C is the covariance matrix of the LSS tracers.
The residual B-mode power can we computed using \refeq{Bres} using an ``effective'' correlation of these combined tracers with gravitational lensing is:
\be
\rho^{2} = \sum_{i,j}\frac{C^{\kappa i}~(C^{-1})_{ij}~C^{\kappa j}}{C^{\kappa\kappa}}.
\label{eqn:rho-combined}
\ee

Note that the gain in adding a new tracer is not only proportional to its correlation with the CMB lensing but it also depends on how much it is correlated with the already used set of tracers.
\reffig{kernel} show the different kernels as a function of redshift computed using the models and parameters described in \refsec{model}.
Galaxies clustering surveys can only reconstruct the low-z portion of the lensing kernel as can be seen from the LSST,DES and DESI curves.
However given the significant low noise of this measurement they can still help the delensing of the CMB.
Furthermore, the low-redshift tail of the lensing kernel is also the one affected by large scale structures non-linearities. They can indeed be extremely helpful if future CMB surveys will try to remove that contribution instead of attempting to model it.
On the other end, CIB and 21 cm surveys are probing to high redshift structures and independently from the model assumed they show a fairly good overlap with the CMB lensing kernel.

@@

\begin{figure}[htbp]
\begin{center}
\includegraphics[scale=1.]{../images/compare_kernel.pdf}
\caption{\textbf{Kernels Comparison}: Comparison of the different kernels as a function of redshift for some of the tracers used in this analysis. The bigger the overlap wit the CMB lensing kernel the better the reconstruction of the lensing potential will be. This will lead to a better delensing. }
\label{fig:kernel}

@@

\section{Lensing potential tracers}
\label{sec:model}
In this section, we describe how the power spectra defined in \refsec{th} are computed and how we characterize different tracers considered in this work.

Large scale structure surveys usually probe the 3D dark matter overdensities as a 2D field projected along the line of sight:

\begin{equation}
\delta^{i}(\nver) = \int_0^{\infty} dz\, W^{i}(z)\delta(\chi(z)\nver,z).
\label{eqn:wkernel}
\end{equation}

where $\delta(\chi(z)\nver,z)$ corresponds to the dark matter overdensity field at a comoving distance $\chi(z)$ and at a redshift $z$ in the angular direction $\nver$.
Using the Limber approximation \cite{limber53} we can compute the power spectra of two large scale structure fields $i,j$ as:
\begin{equation}
C_{\ell}^{ij}= \int_0^{\infty} \frac{dz}{c} \frac{H(z)}{\chi(z)^2}\, W^{i}(z)W^{j}(z)P(k,z).
\label{eqn:wkappa}{}
\end{equation}

In this equation, $H(z)$ is the Hubble factor at redshift $z$, $c$ is the speed of light, $\chi(z)$ is the comoving distance and $P(k,z)$ is the matter power spectrum evaluated at wavenumber $k = \ell/\chi(z)$ and redshift $z$. 
Furthermore $W^{i}(z)$ is the kernel function of the field $i$ defined in \refeq{wkernel}.
We compute these quantities using CAMB and HALOFIT.

\subsection{CMB lensing potential}
\label{sec:kappaCMB}
The first field we compute is the CMB lensing potential.
The lensing kernel $W^{\kappa}$ is:
\begin{equation}
W^{\kappa}(z) = \frac{3\Omega_{\rm m}}{2c}\frac{H_0^2}{H(z)}(1+z)\chi(z)\frac{\chi_*-\chi(z)}{\chi_*},
\end{equation}
%
where $\chi(z)$ is the comoving distance to redshift $z$, $\chi_*$ is the comoving distance to the last-scattering surface at $z_*\simeq 1090$, $\Omega_{\rm m}$ and $H_0$ are the present day values of the Hubble and matter density parameters, respectively.

The CMB lensing potential is \textit{the} field that we need to reconstruct to reverse the effect of large scale structure and delens the CMB.
However, the lensing potential can also be reconstructed from the CMB itself. In that case, we can treat it as a noisy tracer of the true field. Both the CMB lensing fields and its noisy reconstructed counterparts have the same kernel $W^{\kappa}(z)$.
However, when computing the power spectrum of the latter, we need to add a noise component. Given the instrumental noise level and the beam, we can compute the reconstruction noise $N_l^{\kappa \kappa}$, and so
\beq
\cl^{\kappa_\mathrm{rec} \kappa_\mathrm{rec}} = \cl^{\kappa \kappa } +N^{\kappa \kappa}_l
\eeq
In this work, the level of noise is computed assuming an iterative quadratic estimator is used in the reconstruction

\begin{figure}[htbp]
\begin{center}
\includegraphics[scale=1.]{../images/cmb_internal.pdf}
\caption{\textbf{CMB internal delensing:} Correlation factor between CMB internal reconstructed potential and the true lensing potential for different CMB experiments and corresponding noise reconstructions. }
\end{center}
\label{fig:corr-internal}

@@

We can also write the galaxy overdensity field $g(\nver)$ in a given direction on the sky as a line of sight integral of the matter overdensity:
\begin{equation}
g(\nver) = \int_0^{z_*} dz\, W^{g}(z)\delta(\chi(z)\nver,z),
\end{equation}
where the kernel is
\begin{equation}
\label{eqn:wg}
\begin{split}
W^{g}(z) &= \frac{b(z)\frac{dN}{dz}}{\Bigl(\int dz'\,\frac{dN}{dz'}\Bigr)}.
\end{split}
\end{equation}
Here $\frac{dN}{dz}$ is the number of galaxies as a function of redshift observed by the survey while $b(z)$ is the galaxy bias that connects the amplitude of galaxies overdensity to the one of the dark matter at different redshift.
When computing the auto-spectrum of a galaxies density, a shot noise term needs to be taken into account. 
To do so, we add a constant term to the power spectrum equal to the inverse of the number of galaxies per steradians.

Different galaxy surveys in this work are fully characterized by their b(z), $\frac{dN}{dz}$ and the observed galaxies density.
We test the delensing efficiency modeling both current survey like WISE or DES as well as future galaxy probes like DESI and LSST together with 21 cm measurement like SKA.

We defined the WISE redshift distribution following \cite{yan:2013} (see Fig. 4 therein). 
To compute the noise term, we assume that the available sky after masking WISE is around $f_{sky}=0.44$ with 50 million galaxies \cite{ferraro:2015} and that the galaxies density is approximately uniform.
DES is modeled after the DES Science Verification data public release.
For DESI we used the $\frac{dN}{dz}$ in Tab. 2.3 of the DESI Technical Design Report.
LSST is modeled following \cite{schaan:2016} as $\frac{dN}{dz}\propto z^{\alpha}\exp^{-(z/z_{0})^{\beta}}$ with $\alpha = 1.27$, $\beta = 1.02$, and $z_{0}= 0.5$. Furthermore we assume a density of 26 galaxies per arcmin squared.
Finally, we model both the SKA redshift distribution and bias survey following \cite{namikawa:2016a}.

\subsection{Cosmic infrared Background (CIB)}
\label{sec:cib}
Following \cite{Addison:2011se}, we model the CIB power directly as $C_{\ell}^{{\rm CIB}\mbox{-}{\rm CIB}} = 3500 (l/3000)^{-1.25} {\rm Jy^2 / sr}$.
This model provides an accurate fit to several experimental results.
For the cross-spectra with the CMB lensing or other galaxies tracers $C_{\ell}^{{\rm CIB}\mbox{-}j}$, we use the single-SED model of~\cite{hall10}.
It corresponds to the kernel:
\begin{align}\label{eq:12}
W^{\text{CIB}}(z) = b_c\ \frac{\chi^2(z)}{H(z)(1+z)^2}\ e^{-\frac{(z-z_c)^2}{2\sigma^2_z} } f_{\nu(1+z)},
\end {align}
for
\begin{equation}
f_{\nu} = 
\begin{cases}
\Big( e^{\frac{h\nu}{kT}} - 1 \Big)^{-1} \nu^{\beta+3} & (\nu \leq v^{\prime}) \\ \Big( e^{\frac{h\nu^{\prime}}{kT}} - 1 \Big)^{-1} \nu^{\prime \beta+3} \Big( \frac{\nu}{\nu^{\prime}} \Big)^{-\alpha} & (\nu > v^{\prime})
\end{cases}
\end{equation}

We place the peak of the CIB emissivity at redshift $z_c = 2$ with a broad redshift kernel of width $\sigma_z = 2$. Also $T$ = 34K and $\nu^{\prime} \approx$ 4955 GHz.
